---
title: "Caudate Bootstrapping Analysis"
author: "Siddharth Reed"
date: "`r Sys.Date()`"
output: 
    html_document:
        keep_md: true
        toc: true
        toc_float: true
        theme: paper
        code_folding: hide
        df_print: paged
---
<style type="text/css">
    .main-container {
        max-width: 1800px !important;
        margin-left: auto;
        margin-right: auto;
    }
</style>

# Dependencies

```{r knitr}
library(knitr)
knitr::opts_chunk$set(
    dev=c('png', 'pdf', 'tiff'),
    dpi=300
)
```
Set up file locations + load dependencies
```{r dependencies, results=FALSE, message=FALSE, warning=FALSE}
library(here)
here::i_am('notebooks/analysis.Rmd')
BASE_DIR <- here()
SCRIPT_DIR <- here('scripts')
source(file.path(SCRIPT_DIR, 'locations.R'))
source(file.path(SCRIPT_DIR, 'plot_utils.R'))
source(file.path(SCRIPT_DIR, 'bootstrap_utils.R'))
library(tidyverse)
library(magrittr)
library(ggplot2)
library(devtools)
# Compute DEG results in parallel on all bootstraps for each chunk (in serial)
library(furrr)
library(glue)
message(glue('Using {length(availableWorkers())} threads'))
plan(multisession, workers=length(availableWorkers()))
```
Load sample data + expression
```{r load_external_data, message=FALSE}
# Load sample metadata
sample.metadata <- 
    load_sample_metadata()  # in bootstrap.utils.R
# Load counts
counts.matrix <- 
    load_counts()  # in bootstrap.utils.R
# Load list of parameters for each model
MODEL.TISSUE <- 'CAU'
model.parameters <- 
    load_model_specifications(
        sample.metadata=sample.metadata,
        output_dir=file.path(BOOTSTRAP_DIR, MODEL.TISSUE),
    )
```
Set convenient variable specific for this model
```{r set_params}
MODEL.NAME                 <- 'SVA.only'
TAF1_ENSEMBLID             <- 'ENSG00000147133'
this.model.parameters      <- model.parameters %>% filter(model.name == MODEL.NAME)
MODEL.FILENAME             <- this.model.parameters$model.filename[[1]]
# MODEL.NAME                 <- this.model.parameters$model.name[[1]]
EFFECT.VARIABLE            <- this.model.parameters$effect.variable[[1]]
INCLUDE.SVS                <- this.model.parameters$include.SVs[[1]]
COVARIATE.VARIABLES        <- this.model.parameters$covariate.variables[[1]] %>% str_split(',')
FILTER.VARIABLE            <- this.model.parameters$filter.variable[[1]]
CPM.CUTOFF                 <- this.model.parameters$CPM.cutoff[[1]]
MIN.FRAC.SAMPLES.EXPRESSED <- this.model.parameters$min.frac.samples.expressed[[1]]
INTERCEPT                  <- this.model.parameters$intercept[[1]]
N.SAMPLES.PER.GROUP        <- this.model.parameters$n.samples.per.group[[1]]
SEED                       <- this.model.parameters$seed[[1]]
INFO.FILEPATH              <- this.model.parameters$info.filepath[[1]]
MODEL.DIR                  <- this.model.parameters$model.dir[[1]]
LIST.DIR                   <- this.model.parameters$list.dir[[1]]
RESULTS.DIR                <- this.model.parameters$results.dir[[1]]
```

# Load Results with All Samples

Load DESeq2 results using all samples as input
```{r full_model}
full.model.results <- 
    file.path(MODEL.DIR, 'full.model.results.tsv') %>% 
    read_tsv(show_col_types=FALSE) %>%
    add_column(model.name=MODEL.NAME) %>% 
    group_by(contrast) %>% 
    mutate(fdr=p.adjust(pvalue, method='BH')) %>%
    ungroup()
```
Print pvalue distributions
```{r full_model_pvalues}
full.model.results %>%
    group_by(model.name, contrast) %>%
    summarize(
        as_tibble_row(summary(pvalue)),
        as_tibble_row(summary(fdr))
    )
```

# Summarize Bootstrap Results

If you load all bootstrap results this will result in 
```
~20,000 genes * 50,000 bootstraps * 3 stats * C contrasts
~= 4,000,000,000 row X 5 column dataframe per contrast
```

stats are (`pvalue`, `lfc`, `lfcSE`). 
Thankfully only care about main contrast (Genotype XDP / Control).
Instead we will summarize results for each chunk and aggregate summary stats, which leads to a ~20,000 row x 1+ column dataframe where each column is a summary stat or pattern frequency for that gene across all 50,000 bootstraps.
```{r summarize_all_chunks, message=FALSE}
bootstrap.summary.df <- 
    model.parameters %>%
    summarize_all_contrasts(
        model.names=c('SVA.only'),
        contrasts=c('Genotype-XDP-Control'),
        force_redo=FALSE
    )
```
Now print frequencies  distributions for different patterns
```{r bootstrap_pattern_frequency_distribution_table, message=FALSE}
bootstrap.summary.df %>% 
    select(
        -c(
            EnsemblID,
            model.params
        )
    ) %>% 
    pivot_longer(
        -c(
            # EnsemblID,
            model.name,
            contrast
        ),
        names_to='stat',
        values_to='value'
    ) %>%
    filter(str_detect(stat, 'freq.')) %>% 
    group_by(
        model.name,
        contrast,
        stat
    ) %>% 
    summarize(stats=as_tibble_row(summary(value, na.rm=TRUE))) %>% 
    ungroup() %>% 
    unnest(stats) %>%
    separate_wider_delim(
        'stat',
        delim='.',
        names=
            c(
                'stat',
                'significance',
                'direction'
            )
    ) %>%  
    filter(
        significance != 'any',
        direction != 'either',
    ) %>% 
    arrange(
        contrast,
        desc(direction),
        desc(significance)
    ) %>% 
    select(
        -c(
            model.name,
            stat,
            Min.,
        )
    )
```

# Check TAF1 expression

Check expression of TAF1 in the full model (all samples) and across bootstraps.
```{r check_TAF1_full}
# bootstrap.summary.df %>% count(model.name, contrast, model.params, EnsemblID)
# TAF1 results using all samples
full.model.results %>%
    filter(EnsemblID == TAF1_ENSEMBLID) %>% 
    select(-c(variable, numerator, denominator))
```
```{r check_TAF1_bs}
bootstrap.summary.df %>% 
    filter(EnsemblID == TAF1_ENSEMBLID) %>% 
    select(
        -c(
            EnsemblID,
            model.params
        )
    ) %>% 
    pivot_longer(
        -c(
            model.name,
            contrast
        ),
        names_to='stat',
        values_to='value'
    ) %>%
    filter(str_detect(stat, 'freq.')) %>% 
    arrange(model.name, contrast, stat)
```

# Bootstrap Reproducibility Distributions {.tabset}

```{r make_distribution_theme}
theme_obj <- 
    theme(
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.background=element_blank(),
        axis.line=
            element_line(
                color="black",
                linewidth=1/2.13
            ),
        axis.ticks=
            element_line(
                color="black",
                linewidth=1/2.13
            ),
        axis.ticks.length=unit(3, "pt"),
        axis.title=
            element_text(
                family="sans",
                face="bold",
                color="black",
                size=10
            ),
        axis.text=
            element_text(
                family="sans",
                face="bold",
                color="black",
                size=8
            ),
        legend.position='top',
        strip.text=
            element_text(
                size=10,
                face='bold'
           ),
    )
```
```{r distribution_plot_df}
bootstrap.frequency.df <- 
    bootstrap.summary.df %>% 
    select(
        c(
            model.name,
            contrast,
            EnsemblID,
            starts_with('freq.')
        )
    ) %>%
    pivot_longer(
        -c(
            model.name,
            contrast,
            EnsemblID
        ),
        names_to='stat',
        values_to='frequency'
    ) %>%
    separate_wider_delim(
        'stat',
        delim='.',
        names=
            c(
                'tmp',
                'significance',
                'direction'
            )
    ) %>%  
    select(-c(tmp))
bootstrap.frequency.df
```

## Distribution across all bootstraps

Distribution of frequency of each expression pattern observed across all genes 
```{r distribution_plots, message=FALSE}
bootstrap.frequency.df %>% 
    filter(
        # significance %in% c('fdr', 'nominal'),
        # direction != 'either    '
        significance != 'attempts',
        !(significance == 'any' & direction == 'either')
    ) %>% 
    plot_distribution(
        x_var='frequency',
        facet_col='direction',
        facet_row='significance',
        scales='free_y',
        expand=c(0.00, 0.00, 0.00, 0.00),
        # label.size=3,
        label.y=1000,
        # alpha=0.8,
        # linewidth=0.75,
        binwidth=0.02,
        # x_off=0.05, # 1725,
        x_off=-0.02, # 1725,
        theme_obj=theme_obj
    ) +
    scale_x_continuous(limits=c(-0.05, 1))
```

## Distributions for subsets of bootstraps

Distribution of frequency of each expression pattern observed across all genes in different subsets of bootstraps, shows that frequency results are consistent between subsets.
```{r subsampled_distribution_plots, eval=TRUE, message=FALSE}
CONTRAST='Genotype-XDP-Control'
subsample.frequency.df <- 
    file.path(RESULTS.DIR, CONTRAST) %>% 
    subsample_bootstraps(
        model_name=MODEL.NAME,
        contrast_name=CONTRAST,
        n_subsamples=5,
        seed=9
    )
subsample.frequency.df %>% 
    filter(
        # significance != 'attempts',
        # direction != 'either    '
        # !(significance == 'any' & direction == 'either')
        significance %in% c('nominal', 'fdr'),
        direction %in% c('down', 'up')
    ) %>% 
    plot_distribution(
        x_var='frequency',
        fill_var='chunk_id', 
        # facet_group='expression_pattern',
        facet_col='direction',
        facet_row='significance',
        scales='free_y',
        alpha=0.5,
        x_off=0.015,
        label.y=3000,
        theme_obj=theme_obj
    ) +
    scale_x_continuous(
        limits=c(-0.05, NA),
        expand=c(0.00, 0.00, 0.00, 0.00)
    )
```

# Bootstrap Failures

Distribution of bootstrap failure rate over all bootstraps across genes.
Failure rate is the number of bootstraps where DESeq2 returns `NA` for the fold-change.
```{r bootstrap_failure_rate}
bootstrap.summary.df %>% 
    dplyr::select(
        c(
            model.name,
            contrast,
            EnsemblID,
            n.any.either,
            n.attempts.either
        )
    ) %>%
    dplyr::rename('n.successful.bootstraps'=n.any.either) %>% 
    mutate(freq.successful.bootstraps=n.successful.bootstraps / n.attempts.either) %>%
    # filter(freq.successful.bootstraps == 1)
    ggplot(aes(x=freq.successful.bootstraps)) +
    geom_histogram(
        aes(y=after_stat(count)),
        binwidth=0.02,
        position='identity'
    ) + 
    labs(y='# of Genes') +
    scale_y_continuous(
        limits=c(0, NA),
        expand=c(0.00, 0.00, 0.00, 0.00)
    ) +
    facet_grid(
        cols=vars(model.name),
        rows=vars(contrast),
        scales='fixed'
    ) + 
    theme_obj
```

# Bootstrap Reproducibility vs DESeq2 Specificity {.tabset}

```{r join_data}
joint.results.df <- 
    join_full_and_bootstrap_results(
        bootstrap.summary.df,
        full.model.results
    )
```
```{r prep_plot_data}
plot.df <- 
    joint.results.df %>%
    # unite('expression.pattern', c(significance, direction), sep='.', remove=TRUE) %>% 
    mutate(
        full.model.direction=
            case_when(
                log2FoldChange > 0 & fdr < 0.01 ~ 'Up in Full Model',
                log2FoldChange < 0 & fdr < 0.01 ~ 'Down in Full Model',
                TRUE ~ 'N.S'

            )
    ) %>% 
    pivot_longer(
        c(pvalue, fdr), 
        names_to='full.model.significance',
        values_to='full.model.pvalue'
    ) %>% 
    filter(contrast == 'Genotype-XDP-Control') %>% 
    filter(full.model.significance == 'fdr')
```
```{r make_theme}
theme_obj <- 
    theme(
        panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.background=element_blank(),
        axis.line=
            element_line(
                color="black",
                linewidth=1/2.13
            ),
        axis.ticks=
            element_line(
                color="black",
                linewidth=1/2.13
            ),
        axis.ticks.length=unit(3, "pt"),
        axis.title=
            element_text(
                family="sans",
                face="bold",
                color="black",
                size=10
            ),
        axis.text=
            element_text(
                family="sans",
                face="bold",
                color="black",
                size=8
            )
    )
```

## All expresion patterns

```{r bs_vs_deseq_all, message=FALSE}
plot.df %>% 
    plot_compare_results(
        full.BH.thresh=0.01,
        color_var='full.model.direction',
    ) +
    theme_obj
```

## Only relevant patterns for publication figure

Only include the 4 relevant panels
```{r bs_vs_deseq_pub, message=FALSE}
plot.df %>% 
    filter(
        significance %in% c('nominal', 'fdr'),
        direction %in% c('down', 'up'),
    ) %>% 
    plot_compare_results(
        full.BH.thresh=0.01,
        color_var='full.model.direction',
    ) +
    theme_obj
```

## Aloy Data vs Bootstrapping

Now using aloy's data instead of the full model results
```{r aloy_data, message=FALSE}
ALOY_LFC_THRESHOLD <- 0.58
aloy.plot.df <- 
    file.path(BASE_DIR, 'DESeq2_alpha01_CAU.csv')  %>%
    read_csv(show_col_types=FALSE) %>% 
    dplyr::rename(
        'EnsemblID'=ENSG,
        'fdr'=padj
    ) %>% 
    add_column(
        model.name='SVA.only',
        contrast='Genotype-XDP-Control'
    ) %>% 
    join_full_and_bootstrap_results(
        bootstrap.summary.df,
        .
    ) %>% 
    # unite('expression.pattern', c(significance, direction), sep='.', remove=TRUE) %>% 
    mutate(
        full.model.direction=
            case_when(
                log2FoldChange >  ALOY_LFC_THRESHOLD & fdr < 0.01 ~ 'Up in Full Model',
                log2FoldChange < -ALOY_LFC_THRESHOLD & fdr < 0.01 ~ 'Down in Full Model',
                TRUE ~ 'N.S'

            )
    ) %>% 
    pivot_longer(
        c(pvalue, fdr), 
        names_to='full.model.significance',
        values_to='full.model.pvalue'
    ) %>% 
    filter(contrast == 'Genotype-XDP-Control') %>% 
    filter(full.model.significance == 'fdr') %>% 
    filter(
        significance %in% c('nominal', 'fdr'),
        direction %in% c('down', 'up'),
    )
```
```{r bs_vs_deseq_aloy, message=FALSE}
aloy.plot.df %>% 
    plot_compare_results(
        full.BH.thresh=0.01,
        color_var='full.model.direction',
        size=1
    ) +
    theme_obj
```
```{r bs_vs_deseq_aloy_fullcolor, message=FALSE}
aloy.plot.df %>% 
    plot_compare_results(
        full.BH.thresh=0.01,
        color_var='log2FoldChange',
        alpha=0.7,
        size=1
    ) +
    theme_obj
```
